An algorithm for the optimization of finite element integration loops

### 2016-03-18 TOMS_Paper/assembly.tex
Integrate feedback from Paul and David

\documentclass[prodmode,acmtecs]{acmsmall} % Aptara syntax

@@

\begin{abstract}
We present an algorithm for the  optimization of a finite element integration loop nest. This algorithm is proven to achieve a locally optimal operation count. In specified circumstances the optimum achieved is global. Extensive numerical experiments demonstrate significant performance improvements over the state of the art in finite element code generation in almost all cases. This validates the algorithm presented here, and illustrate its limitations. 
\end{abstract}

@@

\acmformat{Fabio Luporini, David A. Ham, and Paul H. J. Kelly, 2015. An algorithm for the optimization of finite element integration loops.}

@@

@@

\subsection{Qualitative comparison}
\label{sec:qualitative}
Depending on form and discretization, the relative performance of the two modes, in terms of the operation count, can vary quite dramatically. The presence of derivatives or coefficient functions in the input form tends to increase the size of the geometry tensor, making the traditional quadrature mode preferable for ``complex'' forms. On the other hand, speed-ups from adopting tensor mode can be significant in a wide class of forms in which the geometry tensor remains ``sufficiently small''. The discretization, particularly the relative polynomial order of trial, test, and coefficient functions, also plays a key role in the resulting operation count.

@@

%TODO is "not expanded yet == still contracted" ?

@@

\begin{Def}[Flop-decreasing transformation]
A sequence of transformations reducing the operation count is called flop-decreasing.
\end{Def}

@@

Achieving optimality in polynomial time is not generally feasible, since the $\sigma_{ei}$ sub-expressions can be arbitrarily unstructured. On the other hand, multilinearity suggests a certain degree of regularity for $\alpha_{eij}$ and $\beta_{eik}$. In the following sections, we will elaborate on these observations and formulate an approach that achieves: (i) at least a local optimum in all cases; (ii) global optimality whenever the monomials are ``sufficiently structured''. To this purpose, we will construct:
\begin{itemize}
\item the space of legal transformations impacting the operation count (Sections~\ref{sec:sharing-elimination} -- \ref{sec:completeness})
\item an algorithm to select points in the transformation space (Section~\ref{sec:optimal-synthesis})
\end{itemize}

@@

In this section, we study \textit{sharing elimination}, a transformation that aims to reduce the operation count by removing sharing through the application of expansion, factorization, and generalized code motion. If the objective were reaching optimality and the expressions lacked structure, a transformation of this sort would require solving a large combinatorial problem -- for instance to evaluate the impact of all possible factorizations. Our sharing elimination strategy, instead, exploits the structure inherent in finite element integration expressions to guarantee local optimality\footnote{This requires coordination with other transformations, as discussed in the following sections. For the moment -- and for ease of exposition -- we neglect this aspect.}, with optimality being achieved if stronger preconditions hold. Setting local optimality, instead of optimality, as primary goal is essential to produce simple and computationally efficient algorithms -- two necessary conditions for integration with a compiler.

@@

\subsubsection{Identification and exploitation of structure}
\label{sec:se-rln}
Finite element expressions can be seen as composition of operations between tensors. Often, the optimal implementation strategy for these operations is to be determined out of two alternatives. For instance, consider $J^{-1} \nabla v \cdot J^{-1} \nabla v$, with $J^{-1}$ being the inverse Jacobian matrix for the change of coordinates and $v$ a generic two-dimensional vector. The tensor operation will reduce to the scalar expression $(a v^0_i + b v^1_i) (a v^0_i + b v^1_i) + ...$, in which $v^0_i$ and $v^1_i$ represent components of $v$ that depend on $L_i$. To minimize the operation count for expressions of this kind, we have two options.
\begin{Strategy}
\label{strategy:i}
Eliminating temporal sharing through generalized code motion.
\end{Strategy}
\begin{Strategy}
\label{strategy:ii}
Eliminating spatial sharing first -- through product expansion and factorization -- and temporal sharing afterwards, again through generalized code motion.
\end{Strategy}
In the current example, we observe that, depending on the size of $L_i$, applying Strategy~\ref{strategy:ii} could reduce the operation count since the expression would be recast as $v^0_i v^0_i a a + v^0_i v^1_i (ab + ab) + v^1_i v^1_i c c + ...$ and some hoistable sub-expressions would be exposed. On the other hand, Strategy~\ref{strategy:i} would have no effect as $v$ only depends on a single loop, $L_i$. In general, the choice between the two strategies depends on multiple factors: the loop(s) size, the increase in operation count due to expansion (in Strategy~\ref{strategy:ii}), and the gain in code motion. A second application of Strategy~\ref{strategy:ii} was provided in Figure~\ref{code:multi_loopnest}. These examples motivate the introduction of a particular class of expressions, for which the two strategies assume notable importance.
\begin{Def}[Structured expression]
\label{def:struct-expr}
We say that ``an expression is structured along a loop nest $\Lambda$'' if and only if in order to eliminate the spatial sharing induced by any $s_{\Lambda}$ -- a generic symbol depending on at least one loop in $\Lambda$ -- it is sufficient to factorize all occurrences of $s_{\Lambda}$ in the expression.
\end{Def}
\begin{Prop}
\label{prop:multi-struct}
An expression along a multilinear loop nest is structured.
\end{Prop}
\begin{proof}
This descends directly from Definition~\ref{def:linear-loop} and Definition~\ref{def:multi-linear-loop}, which essentially restrict the number of occurrences of a symbol $s_{\Lambda}$ in a summand to at most 1.
\end{proof}
If $\Lambda$ were an arbitrary loop nest, a given symbol $s_{\Lambda}$ could appear everywhere (e.g., $n$ times in a summand and $m$ times in another summand with $n \neq m$, as argument of a higher level function, in the denominator of a division), thus posing the challenge of finding the factorization that maximizes temporal sharing. If $\Lambda$ is instead a finite element integration loop nest, thanks to Proposition~\ref{prop:multi-struct} the space of flop-decreasing transformations is constructed by ``composition'' of Strategy~\ref{strategy:i} and Strategy~\ref{strategy:ii}, as illustrated in Algorithm~\ref{algo:sharing-elimination}.

Finally, we observe that the $\sigma_{ei}$ sub-expressions can sometimes be considered ``weakly structured''. This happens when a relaxed version of Definition~\ref{def:struct-expr}, in which the factorization of $s_{\Lambda}$ only ``minimizes'' (rather than ``eliminates'') spatial sharing, applies (for instance, in the complex hyperelastic model analyzed in Section~\ref{sec:perf-results}). Weak structure will be exploited by Algorithm~\ref{algo:sharing-elimination} in the attempt to achieve optimality.

@@

\begin{Algo}[Sharing elimination]
\label{algo:sharing-elimination}
\normalfont 
~
\begin{enumerate}
\item Perform a depth-first visit of the expression tree to collect and partition multilinear operands into disjoint sets, $\mathbb{P} = \lbrace P^1, ..., P^{p}\rbrace$. $\mathbb{P}$ is such that all multilinear operands in $P$ share the same set of multilinear symbols $S_P$, whereas there is no sharing across different partitions. For all multilinear operands in $P \in \mathbb{P}$ such that $|P| \leq |S_P|$, apply Strategy~\ref{strategy:i}. \\
\textit{Note: as a consequence of Proposition~\ref{prop:multi-struct}, $|P|$ and $|S_P|$ represent the number of products in the innermost loop induced by $P$ if Strategy~\ref{strategy:i} or Strategy~\ref{strategy:ii} were applied}

\item For each sub-expression $expr$ depending on exactly one linear loop, collect the multilinear symbols and the temporaries produced at step (1). Partition them into disjoint sets, $\mathbb{T} = \lbrace T^1, ..., T^t\rbrace$, such that $T^i$ includes all instances of a given symbol in $expr$. Apply Strategy~\ref{strategy:ii} factorizing the symbols in each $T^i$, provided that this leads to a reduction in operation count; otherwise, apply Strategy~\ref{strategy:i} \\
\textit{Note: the last check ensures the flop-decreasing nature of the transformation. In the cases in which expansion outweighs code motion, Strategy~\ref{strategy:i} is preferred.}\\
\textit{Note: the expansion cost is a function of the products wrapping a symbol (how many of them and their arity), so it can be determined in a tree visit.}

\item Build the \textit{sharing graph} $G = (S, E)$. Each $s \in S$ represents a multilinear symbol or a temporary produced by the previous steps. An edge $(s^i$, $s^j)$ indicates that a product $s^i s^j$ would appear if the sub-expressions including $s^i$ and $s^j$ were expanded.\\
\textit{Note: the following steps will only impact bilinear forms, since otherwise $E = \emptyset$.}

\item Partition $S$ into disjoint sets, $\mathbb{S} = \lbrace S^1, ..., S^n\rbrace$, such that $S^i$ includes all instances of a given symbol $s$ in the expression. Transform $G$ by merging $\lbrace s^1, ..., s^m \rbrace \subset S^i$ into a unique vertex $s$ (taking the union of the edges), provided that factorizing $[s^1, ..., s^m]$ would not cause an increase in operation count.

\item Map $G$ to an Integer Linear Programming (ILP) model for determining how to optimally apply Strategy~\ref{strategy:ii}. The solution is the set of symbols that will be factorized by Strategy~\ref{strategy:ii}. Let $|S| = n$; the ILP model then is as follows:
\begin{gather*}
x_i \text{: a vertex in } S \text{ (1 if a symbol should be factorized, 0 otherwise)}\\
y_{ij} \text{: an edge in } E \text{ (1 if } s^i \text{ is factorized in the product } s^i s^j \text{ , 0 otherwise)}\\
n_i \text{: the number of edges incident to } x_i \\
\begin{align*}
\min \sum_{i=1}^{n} x_i,\ s.t. ~~~~~~~~~&\sum_{j|(i,j) \in E} y_{ij} \leq n_i x_i,\ i = 1, ..., n \\
& y_{ij} + y_{ji} = 1,\ (i, j) \in E
\end{align*}
\phantom{\hspace{6cm}}
\end{gather*}

\item Perform a depth-first visit of the expression tree and, for each yet unhandled or hoisted expression, apply the most profitable between Strategy~\ref{strategy:i} and Strategy~\ref{strategy:ii}. \\
\textit{Note: this pass speculatively assumes that expressions are (weakly) structured along the reduction loop. If the assumption does not hold, the operation count will generally be sub-optimal because only a subset of factorizations and code motion opportunities may eventually be considered.}
\end{enumerate}
\end{Algo}

\begin{Rem*}
\normalfont
Although the primary goal of Algorithm~\ref{algo:sharing-elimination} is operation count minimization within the multilinear loop nest, the enforcement of flop-decreasing transformations (steps (2) and (4)) and the re-scheduling of sub-expressions within outer loops (last step) also attempt to optimize the loop nest ``globally''. We will further elaborate this aspect in Section~\ref{sec:proof}.
\end{Rem*}

\subsubsection{Examples}
\label{sec:se-examples}
Consider again Figure~\ref{code:multi_loopnest_a}. We have $\mathbb{P} = \lbrace P^0, P^1, P^2\rbrace$, with $P^0 = \lbrace b_j\rbrace$, $P^1 = \lbrace c_i\rbrace$, and $P^2 = \lbrace d_i\rbrace$. For all $P^i$, we have $|P^i| = 1 = |S_{P^i}|$, although applying Strategy~\ref{strategy:i} in step (1) has no effect. The sharing graph is $G = (\lbrace b_j, c_i, d_i \rbrace, \lbrace (b_j, c_i), (b_j, d_i) \rbrace$, and $T = \lbrace c_i, d_i \rbrace$. The ILP formulation leads to the code in Figure~\ref{code:multi_loopnest_c}.

In Figure~\ref{code:poisson}, Algorithm~\ref{algo:sharing-elimination} is executed in a realistic scenario, which originates from the bilinear form of a Poisson equation in two dimensions. We observe that $\mathbb{P} = \lbrace P^0, P^1\rbrace$, with $P^0 = \lbrace (z_0 a_{ik} + z_2 b_{ik}), (z_1 a_{ik} + z_3 b_{ik})\rbrace$ and $P^1 = \lbrace (z_0 a_{ij} + z_2 b_{ij}), (z_1 a_{ij} + z_3 b_{ij})\rbrace$. In addition, $|P^i| = |S_{P^i}| = 2$, so Strategy~\ref{strategy:i} is applied to both partitions (step (1)). We then have (step (3)) $G = (\lbrace t^0, t^1, t^2, t^3 \rbrace, \lbrace (t^0, t^2), (t^1, t^3) \rbrace)$. Since there are no more factorization opportunities, the ILP formulation becomes irrelevant.

\begin{figure}[h]\begin{CenteredBox}
{\subfigcapskip = 7pt \subfigure[Normal form]{\label{code:poisson_a}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/poisson_pre_se.code}}}
~~~~~~~~~~
{\subfigcapskip = 7pt \subfigure[After sharing elimination]{\label{code:poisson_b}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/poisson_after_se.code}}}
\end{CenteredBox}\caption{Applying sharing elimination to the bilinear form arising from a Poisson equation in 2D. The operation counts are $E(f(z^0, z^1, ...) + IJK \cdot 18)$ (left) and $E(f(z^0, z^1, ...) + I(J \cdot 6 + K \cdot 9 + JK \cdot 4)$ (right), with $f(z^0, z^1, ...)$ representing the operation count for evaluating $z^0, z^1, ...$, and common sub-expressions being counted once. Note that the synthesis in Figure~\ref{code:poisson_b} reaches optimality apart from the pathological case  $I, J, K = 1$.}\label{code:poisson}\end{figure}

@@

Pre-evaluation can be seen as the generalization of tensor contraction (Section~\ref{sec:tc}) to a wider class of sub-expressions. We know that multilinear forms can be seen as sums of monomials, each monomial being an integral over the equation domain of products (of derivatives) of functions from discrete spaces. A monomial can always be reduced to the product between a ``reference'' and a ``geometry'' tensors. In our model, a reference tensor is simply represented by one or more sub-expressions independent of $L_e$, exposed after particular transformations of the expression tree. This leads to the following algorithm.

@@

The pre-evaluation of a monomial introduces some critical issues:
\begin{enumerate}
\item Depending on the complexity of a monomial, a certain number $t$ of temporary variables is required if pre-evaluation is performed. Such temporary variables are actually $n$-dimensional arrays of size $S$, with $n$ and $S$ being, respectively, the arity and the extent (iteration space size) of the multilinear loop nest (e.g., $n=2$ and $S = J K$ in the case of bilinear forms). For certain values of ${<}t, n, S{>}$, pre-evaluation will shift the loop nest from a CPU-bound to a memory-bound regime, which may be counter-productive for actual execution time.
\item The transformations exposing $[L_i, L_j, L_k]$-dependent terms increase the arithmetic complexity of the expression (e.g., expansion tends to increase the operation count). This could outweigh the gain due to pre-evaluation.
\item A strategy for coordinating sharing elimination and pre-evaluation is needed. We observe that sharing elimination inhibits pre-evaluation, whereas pre-evaluation could expose further sharing elimination opportunities.
\end{enumerate}

We expand on point 1) in the next section, while we address points 2) and 3) in Section~\ref{sec:optimal-synthesis}.

@@

\subsection{Limitation in the optimization space considered}
\label{sec:completeness}
We defined sharing elimination and pre-evaluation as high level transformations on top of basic operators such as code motion and factorization. Factorization addresses \textit{spatial redundancy}. The presence of spatial redundancy means that some operations are needlessly executed at two points in an expression. Code motion and reduction pre-evaluation, on the other hand, target \textit{temporal redundancy}; that is, the needless execution of the same operation with the same operands at two points in the loop nest. We remark that sharing elimination and pre-evaluation aim to minimize spatial and temporal redundancy relying on \textit{a specific set of operators}. Theoretically, lower operation counts may be found by exploiting other domain-specific properties, such as redundancies in basis functions. Constraint~\ref{const:Le} also limits the applicability of code motion, which further reduced the transformation space to be analyzed.

@@

@@

@@

Experiments were run on a single core of an Intel I7-2600 (Sandy Bridge) CPU, running at 3.4GHz, 32KB L1 cache (private), 256KB L2 cache (private) and 8MB L3 cache (shared). The Intel Turbo Boost and Intel Speed Step technologies were disabled. The Intel \texttt{icc 15.2} compiler was used. The compilation flags used were \texttt{-O3, -xHost}.

@@

The values that we report are the average of three runs with ``warm cache'' (no code generation time, no compilation time). They include the cost of both local assembly and matrix insertion, with the latter minimized through the choice of a mesh (details below) small enough to fit the L3 cache of the CPU.

@@

\paragraph{Autovectorization}
We discretized our problems such that the inner loop sizes would be a multiple of the machine vector length. This ensured autovectorization in the majority of code variants\footnote{The compilation flag \texttt{xHost} helped the Intel compiler generating efficient vector code. We verified the vectorization of inner loops by looking at both compiler reports and assembly code.}. The biggest exception is \texttt{quad}, due to the presence of indirection arrays in the generated code. In \texttt{tens}, loop nests are fully unrolled, so the standard loop vectorization is not feasible; the compiler reports suggest, however, that block vectorization (\cite{SLP-vect}) is often triggered. In \texttt{ufls}, \texttt{cfO1}, and \texttt{cfO2} the iteration spaces have identical structure, with loop vectorization being regularly applied.

@@

The sporadic slow-downs (or marginal improvements) exhibited by \texttt{ufls} are imputable to the presence of sharing in the generated code.

@@

\section{Conclusions}
\label{sec:conclusions}
We have developed transformation algorithms and a theory of optimality for finite element integration. The implementation was carried out in COFFEE, a compiler fully integrated with the Firedrake framework, and therefore made publicly available. Performance results suggest the effectiveness of our methodology. There are still several open problems. Among them, a classification of forms for which optimality can be reached and a generalization of the methodology to other loop nests, for instance those arising in spectral methods.

>>>>

\documentclass[prodmode,acmtoms]{acmsmall} % Aptara syntax

@@

\begin{abstract}
We present an algorithm for the  optimization of a class of finite element integration loop nests. This algorithm is proven to achieve a locally optimal operation count. In specified circumstances the optimum achieved is global. Extensive numerical experiments demonstrate significant performance improvements over the state of the art in finite element code generation in almost all cases. This validates the algorithm presented here, and illustrate its limitations. 
\end{abstract}

@@

\acmformat{Fabio Luporini, David A. Ham, and Paul H. J. Kelly, 2016. An algorithm for the optimization of finite element integration loops.}

@@

After reviewing basic concepts in finite element integration, we introduce in Section~\ref{sec:lnopt} a set of definitions mapping mathematical properties to the level of loop nests. This step is essential for understanding the two algorithms -- sharing elimination (Section~\ref{sec:sharing-elimination}) and pre-evaluation (Section~\ref{sec:pre-evaluation}) -- through which we construct the space of legal transformations. The main ``transformation algorithm'' in Section~\ref{sec:optimal-synthesis} delivers the local optimality claim by using a cost model to ``coordinate'' the application of sharing elimination and pre-evaluation. We elaborate on the correctness of the methodology in Section~\ref{sec:proof}. The numerical experiments are showed in Section~\ref{sec:perf-results}. We conclude discussing the limitations of the algorithms presented and future work.

@@

\subsection{Qualitative comparison}
\label{sec:qualitative}
Depending on form and discretization, the relative performance of the two modes, in terms of the operation count, can vary quite dramatically. The presence of derivatives or coefficient functions in the input form increases the rank of the geometry tensor, making the traditional quadrature mode preferable for ``complex'' forms. On the other hand, speed-ups from adopting tensor mode can be significant in a wide class of forms in which the geometry tensor remains ``sufficiently small''. The discretization, particularly the relative polynomial order of trial, test, and coefficient functions, also plays a key role in the resulting operation count.

@@

@@

\begin{Def}[Flop-decreasing transformation]
A transformation which reduces the operation count is called flop-decreasing.
\end{Def}

@@

Achieving optimality in polynomial time is not generally feasible, since the $\sigma_{ei}$ sub-expressions can be arbitrarily unstructured. On the other hand, multilinearity suggests a certain degree of regularity for $\alpha_{eij}$ and $\beta_{eik}$. In the following sections, we will elaborate on these observations and formulate an approach that achieves: (i) at least a local optimum in all cases; (ii) global optimality whenever the monomials are ``sufficiently structured''. To this purpose, we will construct:
\begin{itemize}
\item the space of legal transformations impacting the operation count (Sections~\ref{sec:sharing-elimination} -- \ref{sec:mem-const})
\item an algorithm to select points in the transformation space (Section~\ref{sec:optimal-synthesis})
\end{itemize}

@@

In this section, we study \textit{sharing elimination}, a transformation that aims to reduce the operation count by removing sharing through the application of expansion, factorization, and generalized code motion. If the objective were reaching optimality and the expressions lacked structure, a transformation of this sort would require solving a large combinatorial problem -- for instance to evaluate the impact of all possible factorizations. Our sharing elimination strategy, instead, exploits the structure inherent in finite element integration expressions to guarantee, after coordination with other transformations (an aspect which we discuss in the following sections), local optimality, with optimality being achieved if stronger preconditions hold. Setting local optimality, rather than optimality, as primary goal is essential to produce simple and computationally efficient algorithms -- two necessary conditions for integration with a compiler.

@@

\subsubsection{Identification and exploitation of structure}
\label{sec:se-rln}
Finite element expressions can be seen as composition of operations between tensors. Often, the optimal implementation strategy for these operations is to be determined out of two alternatives. For instance, consider $J^{-1} \nabla v \cdot J^{-1} \nabla v$, with $J^{-1}$ being the inverse Jacobian matrix for the change of coordinates and $v$ a generic two-dimensional vector. The tensor operation will reduce to the scalar expression $(a v^0_i + b v^1_i) (a v^0_i + b v^1_i) + ...$, in which $v^0_i$ and $v^1_i$ represent components of $v$ that depend on $L_i$. To minimize the operation count for expressions of this kind, we have two options:
\begin{Strategy}
\label{strategy:i}
Eliminating temporal sharing through generalized code motion.
\end{Strategy}
\begin{Strategy}
\label{strategy:ii}
Eliminating spatial sharing first -- through product expansion and factorization -- and temporal sharing afterwards, again through generalized code motion.
\end{Strategy}
In the current example, we observe that, depending on the size of $L_i$, applying Strategy~\ref{strategy:ii} could reduce the operation count since the expression would be recast as $v^0_i v^0_i a a + v^0_i v^1_i (ab + ab) + v^1_i v^1_i c c + ...$ and some hoistable sub-expressions would be exposed. On the other hand, Strategy~\ref{strategy:i} would have no effect as $v$ only depends on a single loop, $L_i$. In general, the choice between the two strategies depends on multiple factors: the loop(s) size, the increase in operation count due to expansion (in Strategy~\ref{strategy:ii}), and the gain in code motion. A second application of Strategy~\ref{strategy:ii} was provided in Figure~\ref{code:multi_loopnest}. These examples motivate the introduction of a particular class of expressions, for which the two strategies assume notable importance.
\begin{Def}[Structured expression]
\label{def:struct-expr}
We say that ``an expression is structured along a loop nest $\Lambda$'' if and only if in order to eliminate the spatial sharing induced by any $s_{\Lambda}$ -- a generic symbol depending on at least one loop in $\Lambda$ -- it is sufficient to factorize all occurrences of $s_{\Lambda}$ in the expression.
\end{Def}
\begin{Prop}
\label{prop:multi-struct}
An expression along a multilinear loop nest is structured.
\end{Prop}
\begin{proof}
This follows directly from Definition~\ref{def:linear-loop} and Definition~\ref{def:multi-linear-loop}, which essentially restrict the number of occurrences of a symbol $s_{\Lambda}$ in a summand to at most 1.
\end{proof}
If $\Lambda$ were an arbitrary loop nest, a given symbol $s_{\Lambda}$ could appear everywhere (e.g., $n$ times in a summand and $m$ times in another summand with $n \neq m$, as argument of a higher level function, in the denominator of a division), thus posing the challenge of finding the factorization that maximizes temporal sharing. If $\Lambda$ is instead a finite element integration loop nest, thanks to Proposition~\ref{prop:multi-struct} the space of flop-decreasing transformations is constructed by ``composition'' of Strategy~\ref{strategy:i} and Strategy~\ref{strategy:ii}, as illustrated in Algorithm~\ref{algo:sharing-elimination}.

Finally, we observe that the $\sigma_{ei}$ sub-expressions can sometimes be considered ``weakly structured''. This happens when a relaxed version of Definition~\ref{def:struct-expr} applies, in which the factorization of $s_{\Lambda}$ only ``minimizes'' (rather than ``eliminates'') spatial sharing (for instance, in the complex hyperelastic model analyzed in Section~\ref{sec:perf-results}). Weak structure will be exploited by Algorithm~\ref{algo:sharing-elimination} in the attempt to achieve optimality.

@@

\noindent\rule[0.01ex]{\linewidth}{0.7pt}

\begin{Algo}[Sharing elimination]
\label{algo:sharing-elimination}
\normalfont 
The input of the algorithm is a tree representation a finite element integration loop nest.
\begin{enumerate}
\item Perform a depth-first visit of the loop tree to collect and partition multilinear operands into disjoint sets, $\mathbb{P} = \lbrace P_1, ..., P_p\rbrace$. $\mathbb{P}$ is such that all multilinear operands in each $P \in \mathbb{P}$ share the same set of multilinear symbols $S_P$, whereas there is no sharing across different partitions. For all multilinear operands in $P \in \mathbb{P}$ such that $|P| \leq |S_P|$, apply Strategy~\ref{strategy:i}. \\
\textit{Note: as a consequence of Proposition~\ref{prop:multi-struct}, $|P|$ and $|S_P|$ represent the number of products in the innermost loop induced by $P$ if Strategy~\ref{strategy:i} or Strategy~\ref{strategy:ii} were applied}

\item For each sub-expression $\operatorname{expr}$ depending on exactly one linear loop, collect the multilinear symbols and the temporaries produced at step (1). Partition them into disjoint sets, $\mathbb{T} = \lbrace T_1, ..., T_t\rbrace$, such that $T_i$ includes all instances of a given symbol in $\operatorname{expr}$. Apply Strategy~\ref{strategy:ii} factorizing the symbols in each $T_i$, provided that this leads to a reduction in operation count; otherwise, apply Strategy~\ref{strategy:i} \\
\textit{Note: the last check ensures the flop-decreasing nature of the transformation. In the cases in which expansion outweighs code motion, Strategy~\ref{strategy:i} is preferred.}\\
\textit{Note: the expansion cost is a function of the products wrapping a symbol (how many of them and their arity), so it can be determined in a tree visit.}

\item Build the \textit{sharing graph} $G = (S, E)$. Each $s \in S$ represents a multilinear symbol or a temporary produced by the previous steps. An edge $(s_i$, $s_j)$ indicates that a product $s_i s_j$ would appear if the sub-expressions including $s_i$ and $s_j$ were expanded.\\
\textit{Note: the following steps will only impact bilinear forms, since otherwise $E = \emptyset$.}

\item Partition $S$ into disjoint sets, $\mathbb{S} = \lbrace S_1, ..., S_n\rbrace$, such that $S_i$ includes all instances of a given symbol $s$ in the expression. Transform $G$ by merging $\lbrace s_1, ..., s_m \rbrace \subset S_i$ into a unique vertex $s$ (taking the union of the edges), provided that factorizing $[s_1, ..., s_m]$ would not cause an increase in operation count.

\item Map $G$ to an Integer Linear Programming (ILP) model for determining how to optimally apply Strategy~\ref{strategy:ii}. The solution is the set of symbols that will be factorized by Strategy~\ref{strategy:ii}. Let $|S| = n$; the ILP model then is as follows:
\begin{gather*}
x_i \text{: a vertex in } S \text{ (1 if a symbol should be factorized, 0 otherwise)}\\
y_{ij} \text{: an edge in } E \text{ (1 if } s_i \text{ is factorized in the product } s_i s_j \text{ , 0 otherwise)}\\
n_i \text{: the number of edges incident to } x_i \\
\begin{align*}
\min \sum_{i=1}^{n} x_i,\ s.t. ~~~~~~~~~&\sum_{j|(i,j) \in E} y_{ij} \leq n_i x_i,\ i = 1, ..., n \\
& y_{ij} + y_{ji} = 1,\ (i, j) \in E
\end{align*}
\phantom{\hspace{6cm}}
\end{gather*}

\item Perform a depth-first visit of the loop tree and, for each yet unhandled or hoisted expression, apply the most profitable between Strategy~\ref{strategy:i} and Strategy~\ref{strategy:ii}. \\
\textit{Note: this pass speculatively assumes that expressions are (weakly) structured along the reduction loop. If the assumption does not hold, the operation count will generally be sub-optimal because only a subset of factorizations and code motion opportunities may eventually be considered.}
\end{enumerate}
\end{Algo}

\noindent\rule[1.0ex]{\linewidth}{0.7pt}

Although the primary goal of Algorithm~\ref{algo:sharing-elimination} is operation count minimization within the multilinear loop nest, the enforcement of flop-decreasing transformations (steps (2) and (4)) and the re-scheduling of sub-expressions within outer loops (last step) also attempt to optimize the loop nest ``globally''. We will further elaborate this aspect in Section~\ref{sec:proof}.

\subsubsection{Examples}
\label{sec:se-examples}
Consider again Figure~\ref{code:multi_loopnest_a}. We have $\mathbb{P} = \lbrace P_0, P_1, P_2\rbrace$, with $P_0 = \lbrace b_j\rbrace$, $P_1 = \lbrace c_i\rbrace$, and $P_2 = \lbrace d_i\rbrace$. For all $P_i$, we have $|P_i| = 1 = |S_{P_i}|$, although applying Strategy~\ref{strategy:i} in step (1) has no effect. The sharing graph is $G = (\lbrace b_j, c_i, d_i \rbrace, \lbrace (b_j, c_i), (b_j, d_i) \rbrace$, and $T = \lbrace c_i, d_i \rbrace$. The ILP formulation leads to the code in Figure~\ref{code:multi_loopnest_c}.

In Figure~\ref{code:poisson}, Algorithm~\ref{algo:sharing-elimination} is executed in a very simple realistic scenario, which originates from the bilinear form of a Poisson equation in two dimensions. We observe that $\mathbb{P} = \lbrace P_0, P_1\rbrace$, with $P_0 = \lbrace (z_0 a_{ik} + z_2 b_{ik}), (z_1 a_{ik} + z_3 b_{ik})\rbrace$ and $P_1 = \lbrace (z_0 a_{ij} + z_2 b_{ij}), (z_1 a_{ij} + z_3 b_{ij})\rbrace$. In addition, $|P_i| = |S_{P_i}| = 2$, so Strategy~\ref{strategy:i} is applied to both partitions (step (1)). We then have (step (3)) $G = (\lbrace t_0, t_1, t_2, t_3 \rbrace, \lbrace (t_0, t_2), (t_1, t_3) \rbrace)$. Since there are no more factorization opportunities, the ILP formulation becomes irrelevant.

\begin{figure}[h]\begin{CenteredBox}
{\subfigcapskip = 7pt \subfigure[Normal form]{\label{code:poisson_a}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/poisson_pre_se.code}}}
~~~~~~~~~~
{\subfigcapskip = 7pt \subfigure[After sharing elimination]{\label{code:poisson_b}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/poisson_after_se.code}}}
\end{CenteredBox}\caption{Applying sharing elimination to the bilinear form arising from a Poisson equation in 2D. The operation counts are $E(f(z_0, z_1, ...) + IJK \cdot 18)$ (left) and $E(f(z_0, z_1, ...) + I(J \cdot 6 + K \cdot 9 + JK \cdot 4)$ (right), with $f(z_0, z_1, ...)$ representing the operation count for evaluating $z_0, z_1, ...$, and common sub-expressions being counted once. The synthesis in Figure~\ref{code:poisson_b} is globally optimal apart from the pathological case  $I, J, K = 1$.}\label{code:poisson}\end{figure}

@@

Pre-evaluation can be seen as the generalization of tensor contraction (Section~\ref{sec:tc}) to a wider class of sub-expressions. We know that multilinear forms can be seen as sums of monomials, each monomial being an integral over the equation domain of products (of derivatives) of functions from discrete spaces. A monomial can always be reduced to the product between a ``reference'' and a ``geometry'' tensor. In our model, a reference tensor is simply represented by one or more sub-expressions independent of $L_e$, exposed after particular transformations of the expression tree. This leads to the following algorithm.

\noindent\rule[0.01ex]{\linewidth}{0.7pt}

@@

\noindent\rule[1.0ex]{\linewidth}{0.7pt}

The pre-evaluation of a monomial introduces some critical issues:
\begin{enumerate}
\item Depending on the complexity of a monomial, a certain number $t$ of temporary variables is required if pre-evaluation is performed. Such temporary variables are actually $n$-dimensional arrays of size $S$, with $n$ and $S$ being, respectively, the arity and the extent (iteration space size) of the multilinear loop nest (e.g., $n=2$ and $S = J K$ in the case of bilinear forms). For certain values of ${<}t, n, S{>}$, pre-evaluation may dramatically increase the working set, which may be counter-productive for actual execution time.
\item The transformations exposing $[L_i, L_j, L_k]$-dependent terms increase the arithmetic complexity of the expression (e.g., expansion tends to increase the operation count). This could outweigh the gain due to pre-evaluation.
\item A strategy for coordinating sharing elimination and pre-evaluation is needed. We observe that sharing elimination inhibits pre-evaluation, whereas pre-evaluation could expose further sharing elimination opportunities.
\end{enumerate}

We expand on point (1) in the next section, while we address points (2) and (3) in Section~\ref{sec:optimal-synthesis}.

@@

@@

\noindent\rule[0.01ex]{\linewidth}{0.7pt}

@@

\noindent\rule[1.0ex]{\linewidth}{0.7pt}

@@

Experiments were run on a single core of an Intel I7-2600 (Sandy Bridge) CPU, running at 3.4GHz, 32KB L1 cache (private), 256KB L2 cache (private) and 8MB L3 cache (shared). The Intel Turbo Boost and Intel Speed Step technologies were disabled. The Intel \texttt{icc 15.2} compiler was used. The compilation flags used were \texttt{-O3, -xHost}. The compilation flag \texttt{xHost} tells the Intel compiler to generate efficient code for the underlying platform.

@@

The values that we report are the average of three runs with ``warm cache''; that is, with all kernels retrieved directly from the Firedrake's cache, so code generation and compilation times are not counted. The timing includes however the cost of both local assembly and matrix insertion, with the latter minimized through the choice of a mesh (details below) small enough to fit the L3 cache of the CPU.

@@

\paragraph{Autovectorization}
We discretized our problems such that the inner loop sizes would be a multiple of the machine vector length. This ensured autovectorization in the majority of code variants\footnote{We verified the vectorization of inner loops by looking at both compiler reports and assembly code.}. The biggest exception is \texttt{quad}, due to the presence of indirection arrays in the generated code. In \texttt{tens}, loop nests are fully unrolled, so the standard loop vectorization is not feasible; the compiler reports suggest, however, that block vectorization (\cite{SLP-vect}) is often triggered. In \texttt{ufls}, \texttt{cfO1}, and \texttt{cfO2} the iteration spaces have identical structure, with loop vectorization being regularly applied.

@@

The sporadic slow-downs (or marginal improvements) exhibited by \texttt{ufls} can be attributed to the presence of sharing in the generated code.

@@

\section{Limitations and future work}
\label{sec:completeness}
We have defined sharing elimination and pre-evaluation as high level transformations on top of a specific set of rewrite operators, such as code motion and factorization, and we have used them to construct the transformation space. There are three main limitations in this process. Firstly, we do not have a systematic strategy to optimize sub-expressions independent of linear loops; despite we have a mechanism to determine ``how much computation'' should be hoisted at the level of the integration (reduction) loop, it is not clear how to effectively improve the heuristics used at step (6) in Algorithm~\ref{algo:sharing-elimination}. Secondly, lower operation counts may be found by exploiting domain-specific properties, such as redundancies in basis functions; this aspect is completely neglected in this article. Thirdly, with Constraint~\ref{const:Le} we have limited the applicability of code motion, which was essential given the complexity of the problem tackled.
